Learning constraints in spreadsheets and tabular data

### 2016-06-07 Paper/ECAI/tabular.tex
Updated LaTeX

  \begin{subfigure}{.65\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicate headerless tables, dark borders indicate maximal type-consistent blocks.
    Most tables only contain type-consistent columns, however, table~$T_2$ also contains type-consistent row.
    This example was created by combining several Excel sheets from several exercise sessions.
    % I guess you want to save space by having everything in one figure, but it might be useful to have a smaller figure on the first page just to explain this example in a bit more detail (or use a smaller example here); it's quite overwhelming now}
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.35\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}}\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^*, \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^* \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^* , \ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^* \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^* \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\eclookup{\range{T_{4}}{\rangeall}{2}}{\range{T_{4}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^* \\
%
      &~\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
	  % Table 6
      &~\ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}}
  \vspace{-18pt}
  \caption{Constraints present in the running example (left), except \textit{ALLDIFFERENT} (16), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).\\
  $^*$: Not present in original document}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\linewidth]{figures/tabletool.png}
  \caption{This simple table selection tool allows users to select rectangular ranges of cells corresponding (headerless) tables, an explicit orientation can also be provided.}
  \label{fig:visual_tool}
\end{figure}

@@

\subsubsection{Redundancy}
We consider two types of redundancy that we aim to eliminate during search.
As noted in Section~\ref{sec:form:dependencies}, for some constraint templates there are \textit{symmetries} over the arguments that lead to trivially equivalent solutions, for example, $\ecprod{B_R'}{B_1'}{B_2'} \Leftrightarrow \ecprod{B_R'}{B_2'}{B_1'}$.
Such duplicates are avoided by defining a \textit{canonical} form for these constraints.
In practice, we define an arbitrary but fixed block ordering and require those blocks that are interchangeable to adhere to this ordering.
Moreover, among the semantically equivalent constraints product (\ecprod{B_R'}{B_1'}{B_2'}) and division (\eccalc{B_1'}{B_R' / B_2'}) we only added division, as well as only difference (\ecdiff{B_r}{B_1}{B_2}) and not sum (\eccalc{B_1}{B_r + B_2}). However, division and sum could easily be added, as could it be added explicitly in post-processing based on the matching product and difference constraints.

@@

\tias{This would be a good place to say that for aggregates, you do SUBASSIGNMENTS with specialised code...}

@@

\section{Evaluation}\label{sec:evaluation}
\sergey{we need to add a summary of the dataset, avg number of constraints, cells, rows, columns}
In this section we experimentally validate our approach.
We studied three main questions concerning the accuracy (\textit{Q1}), precision (\textit{Q2}) and speed (\textit{Q3}) of our algorithm.

The implementation is illustrated using a case study on the spreadsheet corresponding to the previously introduced example (Figure~\ref{fig:main_example}).
In order to quantify the results and generalize our findings we also evaluate our algorithm on a benchmark of 30 (\samuel{check number}) spreadsheets that we assembled from various sources.

There are three main categories of spreadsheets in the benchmark: spreadsheets from an exercise session teaching Excel \samuel{Including running example?}, spreadsheets from tutorials online and publicly available spreadsheets such as crime statistics or financial reports that demonstrate more real world usage\footnote{(attached to to the submitted paper)}.

In this section we focus on \textit{functional} constraints that could be used in spreadsheets, ignoring the structural constraints \textit{ALLDIFFERENT}, \textit{FOREIGNKEY} and \textit{PERMUTATION}.
All experiments were run using Python 3.5.1 (custom Anaconda environment) on a Macbook Pro, Intel Core i7 2.3~GHz with 16GB RAM.

@@

For this example our primary goal of finding all original constraints is achieved, the recall is $1.0$.
The implementation also returns 6 additional constraints, therefore, our precision is $12/18 = 0.67$, which is rather low compared to other spreadsheets in the benchmark.
The low precision can be explained by the many short vectors which increases the chance for constraints to be true by accident.

\subsection{Benchmark}
All spreadsheets have been converted to CSV files.
Definitions of tables are given (in some harder cases with $\mathit{None}$ values, blocks are defined manually). \samuel{chec}

\samuel{Revisit [what does "implemented" mean here? what's the difference between "implemented" and "algorithm", esp given that we are talking experiments? ] }
\luc{what is not implemented, strange \dots in a paper !!!}
For every spreadsheet a ground-truth has been provided manually, specifying the \textbf{relevant} constraints that are expected to be (re-)discovered.
Not all relevant constraints are currently supported by \sname, therefore, we sometimes consider the subset of \textbf{supported} constraints.
Constraints that are currently outside of the scope of this system, e.g. generic nested mathematical or logical formulas and date-related constraints, are ignored.

\subsubsection*{Q1. How accurate is \sname?}
Our implementation achieves a recall of~$0.97$ for relevant constraints, i.e. it finds~90 of~93 relevant constraints in the benchmark suite.
The recall for supported constraints is $1.0$ (90 of 90).

\subsubsection*{Q2. How precise is \sname?}
Our implementation finds 121 constraints across all spreadsheets, 28 of which are not present in the ground-truth.
Therefore, the calculated precision is \samuel{Check numbers}.
Notable, the average precision per spreadsheet is \samuel{math}.
This gap is explained by the large number of additional constraints produced by few spreadsheets \samuel{More info?}.

Examining the additional constraints more closely reveals that only \samuel{x} of the constraints are spurious, i.e. they hold by accident and can be disproved by adding additional data.
Most constraints are \textit{redundant}, such as duplications or implications of other constraints.
Many redundant constraints are the result of bad spreadsheet design, i.e. copying data in multiple places, and the overlapping role of difference and the sum aggregate.

We consider precision (i.e. avoiding redundant constraints) a secondary goal, focusing primarily on accuracy.
The rationale is that the precision can be increased by post-processing the constraints using entailment or heuristics.

\subsubsection*{Q3. How fast is \sname?}
Concerning the speed of the algorithm we also prioritized accuracy when a trade-off had to be made.
For the \benchsize~spreadsheets in the benchmark our implementation ran in ${\runtotal}s \pm {\runtotalstd}s$.
The execution times vary widely though between spreadsheets, only four spreadsheets taking more than $0.2s$.
Most of the execution time in these cases goes towards searching either aggregate constraints or conditional aggregates.
The search for aggregates will be slow on spreadsheets containing larger groups of numeric data.
For conditional aggregates the number of candidate primary keys (all-different) and numeric vectors determines the running time (e.g. the case study example).

\begin{table}
  \caption{Overview evaluation on benchmark}
  \centering
  \begin{tabular}{lll}
    & \textbf{Total} & \textbf{Spreadsheet (avg)} \\
    \textbf{Relevant constraints} & $93$ & $2.91$ \\
    \textbf{Recall} & $96.77\%$ & $94.27\%$ \\
    \textbf{Precision} & $23.14\%~(28)$ & $8.33\%~(0.88)$ \\
    \textbf{Speed (s)} & ${\runtotal}s \pm {\runtotalstd}s$ & ${\runfile}s \pm {\runfilestd}s$
  \end{tabular}
\end{table}
\samuel{Expand summary}
\sergey{would it make sense to specify the size of a spreadsheet in the number of non-zero cells?}

@@

>>>>

  \begin{subfigure}{.65\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicate headerless tables, dark borders indicate maximal type-consistent blocks.
    Most tables only contain type-consistent columns, however, table~$T_2$ also contains type-consistent rows.
    Due to the location of the header of~$T_2$ we can infer that it consists of row data.
    This example was created by combining several Excel sheets from several exercise sessions.
    % I guess you want to save space by having everything in one figure, but it might be useful to have a smaller figure on the first page just to explain this example in a bit more detail (or use a smaller example here); it's quite overwhelming now}
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.35\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^* , \ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^* \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^*, \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^*, \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^* \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\ecsumc{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{4}}{\rangeall}{4} = PREV(\range{T_{4}}{\rangeall}{4}) + \range{T_{4}}{\rangeall}{2} - \range{T_{4}}{\rangeall}{3} \\
%
    % Table 5
      &~\eclookup{\range{T_{5}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^* \\
%
      &~\eclookup{\range{T_{5}}{\rangeall}{3}}{\range{T_{5}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}
    \end{align*}}
  \vspace{-18pt}
  \caption{Constraints present in the running example (left), except \textit{ALLDIFFERENT} (16), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).\\
  $^*$: Not present in original document}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\linewidth]{figures/tabletool.png}
  \caption{This simple table selection tool allows users to select rectangular ranges of cells corresponding to (headerless) tables.
  An explicit orientation can also be provided.}
  \label{fig:visual_tool}
\end{figure}

@@

\subsubsection{Redundancy}
We consider two types of redundancy that we aim to eliminate during search.
As noted in Section~\ref{sec:form:dependencies}, for some constraint templates there are \textit{symmetries} over the arguments that lead to trivially equivalent solutions, for example, $\ecprod{B_R'}{B_1'}{B_2'} \Leftrightarrow \ecprod{B_R'}{B_2'}{B_1'}$.
Such duplicates are avoided by defining a \textit{canonical} form for these constraints.
In practice, we define an arbitrary but fixed block ordering and require those blocks that are interchangeable to adhere to this ordering.
Moreover, among the semantically equivalent constraints product (\ecprod{B_R'}{B_1'}{B_2'}) and division (\eccalc{B_1'}{B_R' / B_2'}) we only added product, as well as only difference (\ecdiff{B_r}{B_1}{B_2}) and not sum (\eccalc{B_1}{B_r + B_2}). However, division and sum could easily be added, as could it be added explicitly in post-processing based on the matching product and difference constraints.

@@

For \textit{FOREIGNKEY}, \textit{LOOKUP}, aggregates and \textit{PROJECT} we use custom, optimized methods to find subassignments.
Instead of the more generic generate-and-test implementation, we use, for example, caching, maximal range-finding and partial tests.
While conditional aggregates use generate-and-test, they employ similar optimizations in their test method.

@@

\section{Evaluation}\label{sec:evaluation}
In this section we experimentally validate our approach.
We studied three main questions concerning the recall (\textit{Q1}), precision (\textit{Q2}) and speed (\textit{Q3}) of our algorithm.

The implementation is illustrated using a case study on the spreadsheet corresponding to the previously introduced example (Figure~\ref{fig:main_example}).
In order to quantify the results and generalize our findings we also evaluate our algorithm on a benchmark of 34 spreadsheets that we assembled from various sources.

\subsection{Method}
\label{sec:evalualtion:method}
In order to obtain a benchmark of spreadsheets we looked at three sources:
1)~spreadsheets from an exercise session for teaching Excel; 2)~spreadsheets from tutorials online; and 3)~publicly available \textit{data} spreadsheets such as real-world crime statistics or financial reports\footnote{(attached to to the submitted paper)}.
All spreadsheets were converted to CSV and the table definitions are fixed.
Group definitions are only specified if the spreadsheet contains ambiguous \textit{None} values.

For every spreadsheet we specified a ground-truth of constraints, i.e. the \textbf{intended} constraints that are expected to be (re-)discovered.
These were determined using the formulas (if present), context and headers in the original spreadsheets.
We encoded all relevant constraints using our representation (in canonical form).
Not all relevant constraints are currently supported by \sname, therefore, we sometimes consider the subset of \textbf{supported} constraints.
Constraints that are currently outside of the scope of this system are, for example, generic nested mathematical or logical formulas.

We focus on \textit{functional} constraints that could be used in spreadsheets, ignoring the structural constraints \textit{ALLDIFFERENT}, \textit{FOREIGNKEY} and \textit{PERMUTATION}.
We included detected equalities as relevant constraints (in 5~spreadsheets).
All experiments were run using Python 3.5.1 on a Macbook Pro, Intel Core i7 2.3~GHz with 16GB RAM.

@@

For this example our primary goal of finding all original constraints is achieved, the recall (i.e. $\frac{\text{relevant discovered}}{\text{relevant}}$) is~$1.0$.
The implementation also returns 6 additional constraints, therefore, our precision (i.e. $\frac{\text{relevant discovered}}{\text{all discovered}}$) is $12/18 = 0.67$, which is rather low compared to, for example, the spreadsheets in the benchmark.
The low precision can be explained by the many short vectors which increases the chance for constraints to be true by accident.

\subsection{Benchmark}
Table~\ref{tbl:category_overview} gives an overview of the spreadsheets in the different categories and summarizes the results of our experiments.
Notice that for the data category the average number of tables and cells is higher while the number of relevant constraints is lower compared to the other categories.

{\setlength{\tabcolsep}{0.33em}
\begin{table}[]
  \centering
  \caption{Summary of properties and results per spreadsheet category}
  \label{tbl:category_overview}
  \begin{tabularx}{\linewidth}{lccccccc}
 & \multicolumn{2}{c}{\textbf{Exercises} (9)} & \multicolumn{2}{c}{\textbf{Tutorials} (21)} & \multicolumn{2}{c}{\textbf{Data} (4)} \\
 & Overall & Sheet avg & Overall & Sheet avg & Overall & Sheet avg \\ \hline
    \textbf{Tables} & 19 & 2.11 & 48 & 2.29 & 4 & 1 \\ \hline
    \textbf{Cells} & 1231 & 137 & 1889 & 90 & 2320 & 580 \\ \hline
    \textbf{\begin{tabular}[c]{@{}l@{}}Relevant\\[-4pt] Constraints\end{tabular}} & 34 & 3.78 & 52 & 2.48 & 6 & 1.50 \\ \hline \hline
\textbf{Recall} & 0.85 & 0.83 & 0.88 & 0.87 & 1.00 & 1.00 \\ \hline
\textbf{\begin{tabular}[c]{@{}l@{}}Recall\\[-4pt] Supported\end{tabular}} & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 \\ \hline
\textbf{Precision} & 0.97 & 0.98 & 0.70 & 0.91 & 1.00 & 1.00 \\ \hline
\textbf{Speed (s)} & 1.62 & 0.18 & 1.76 & 0.08 & 0.81 & 0.20
\end{tabularx}
\end{table}}

\subsubsection*{Q1. How many relevant constraints are found by \sname?}
The overall recall achieved by \sname is~$0.88~(81/92)$.
Similarly, the overall and average recall per category is high, as shown in Table~\ref{tbl:category_overview}.
The recall for supported constraints is $1.00$, meaning that all supported constraints are found.

\subsubsection*{Q2. How precise is \sname?}
Across all spreadsheets \sname achieves a precision of~$0.79~(81/102)$.
However, the overall precision per category reveals that the precision is much higher for the exercises and data categories ($0.97$ and $1.00$) and much lower for the tutorials category ($0.70$).
Moreover, the average precision per spreadsheet is, for all categories, over $0.90$.
Examining the tutorial spreadsheets confirms that there are a few spreadsheets that have a high number of additional constraints.
A qualitative analysis of these spreadsheets shows that their low precision is caused by duplicate data or multiple ways to calculate the same results.
For these spreadsheets the precision can be increased by post-processing the constraints using entailment (to detect implications) and heuristics (e.g. taking into account the alignment of results).

\subsubsection*{Q3. How fast is \sname?}
The average runtime across all spreadsheets is $0.13s$.
Across categories the average running times fluctuate in the range between $0.09s$ to $0.22s$.
While most spreadsheets can be analyzed very quickly, running times can go up to around $1s$ for others.

Most of the execution time in these cases goes towards searching either aggregate constraints or conditional aggregates. \samuel{More details}
The search for aggregates will be slow on spreadsheets containing larger groups of numeric data.
For conditional aggregates the number of candidate primary keys (all-different) and numeric vectors determines the running time (e.g. the case study example).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\linewidth]{figures/comparison.pdf}
  \caption{Overview of the average properties and performance of spreadsheets within every benchmark category relative to the maximum across all spreadsheets.}
  \label{fig:comparison}
\end{figure}

\subsection{Testing the limits}
On many spreadsheets \sname achieves good results.
However, factors such as incompatible formatting, bad spreadsheet design and rounding issues can lead to decreased recall, many additional constraints and long running times due to an exponential blow-up of candidate constraints.
For example, we tested a spreadsheet (``exps\_tias'') with 6~tables, 531~cells and many duplicate columns across tables.
While we obtained a recall of~$1.0$, the algorithm ran for $14.30s$ and the overall precision was~$0.52$ and only~$0.11$ when ignoring equalities.
This shows the need for a (heuristic) post-processing step to increase precision and further optimizations to decrease running times.
